Factor Analysis on Citations, Using a Combined Latent and Logistic Regression Model

### 2019-10-24 Manuscript for submission/manuscript.tex
polishing

\subsubsection{Synthetic Setting} \label{sec:Syn-set}
%In this subsection, we consider the following $3$ scenarios:
%In this subsection, we consider $3$ scenarios.
%In first scenario, we examine a simple setting, in which every paper in the network has only one topic.
%In second scenario, we study a more complicated situation than the first one where each paper in the network can possibly have more than one topic.
%It is worth noting that different clusters of papers, where each cluster is comprised of papers with same topic, are connected via ad-hoc links.
%In third scenario, we consider a situation
Before specifying the network settings in two scenarios that we want to explore, let us first describe a set of procedures for setting the model parameters, $\alpha^*, F^*, D^*$ and $S^*$ sequentially.
We put astroids in the superscript of parameters to denote that they are ground truth.
Readers can refer the meaning of each parameter in the model in Section \ref{sec:intro} and  \ref{sec:model-form}.

@@

    \item Recall that the binary factor loading matrix $F^*$ encodes the relation between topics and papers
        (i.e., if $i$th paper studies about $k$th topic, we denote $F^*_{ki}=1$, otherwise $F^*_{ki}=0$).
        First, we assume that there are $n$ papers in the network, and $K$ topics are embedded in it.
        Each of them consists of $\frac{n}{K}$ papers.
        This can be expressed in $F^*$ as follows:
        \[
            F^* =
               \begin{bmatrix}
                 \smash[b]{\block{n/K}} \\
                 && \smash[b]{\block{n/K}} \\
                 &&&& \ddots \\
                 &&&&& \block{n/K}
                \end{bmatrix}
              \in R^{K \times n},
        \]
        where each row of $F^*$ has $\frac{n}{K}$ $1$'s and each column has only one $1$.
        Note that the remaining entries of the matrix are filled with zeros.
        Then, we assume that $n_{l}$ papers share $l$ topics and $n_{m}$ papers share $m$ topics ($1 < l < m \leq K$).
        This can be represented in the $F^*$ in following steps:
        $1)$ Pick distinct $n_{l}$ indices randomly from $\{1,2,\dots,n\}$.
        We will denote the set of the indices as $\Omega_{n_l}$.
        $2)$ Choose $n_{m}$ indices from the set $\{1,2,\dots,n\}\setminus \Omega_{n_{l}}$ and denote the set of those indices as $\Omega_{n_{m}}$.
        $3)$ Make the columns of $F^*$ with corresponding indices in set $\Omega_{n_{l}} \cup \Omega_{n_{m}}$ zeros.
        $4)$ We use a notation $f^*_j$ to denote the $j$th column of the matrix $F^*$.
        Fill arbitrary $l$ entries of $f^*_j$ for $j \in \Omega_{n_l}$ with $1$s, and also fill arbitrary $m$ entries of $f^*_j$ for $j \in \Omega_{n_m}$ with $1$s.
        Lastly, we set $F^*=JF^*$ where $J=I_{n}-\frac{1}{n}\mathbbm{1}\mathbbm{1}^T$.

@@

    \item Recall that the positive entries of $S^*$ can characterize the links in the network which cannot be accounted by the common topics.
        In the spirit of this notion, we construct ad-hoc links in a way that they connect the clusters of papers with one topic.
        Specifically, we assume that there are $\frac{|S^*|}{\binom{K}{2}}$ edges between two clusters, where $|S^*|$ denotes the number of non-zero entries of the upper-triangular part of the matrix $S^*$.
        This can be implemented via following steps:
        1) We construct $K$ sets $C^*_1,\dots,C^*_K$ that are defined as follows:
        \begin{align*}
           &C^*_{1} \subseteq \big\{1, 2, \dots, \frac{n}{K} \big\} \setminus \big\{ \Omega_{n_l}
           \cup \Omega_{n_m} \big\} \\
           &C^*_{2} \subseteq \big\{\frac{n}{K}+1, \dots, \frac{2n}{K} \big\} \setminus \big\{ \Omega_{n_l} \cup \Omega_{n_m} \big\}  \\
           & \quad \vdots \\
           &C^*_{K} \subseteq \big\{\frac{(K-1)n}{K}+1, \dots, n \big\} \setminus \big\{ \Omega_{n_l}
           \cup \Omega_{n_m} \big\}
        \end{align*}
        where each of them has arbitrary $\frac{|S^*|}{\binom{K}{2}}$ elements.
        2) Create a set $I_{S^*}$ whose elements are pairs of indices such that
        \[
            I_{S^*} =
            \bigg\{ (i_r,j_r) : i_r \in C^*_p, j_r \in C^*_q, 1 \leq p < q \leq K,   r = 1,2,\dots,\frac{|S^*|}{\binom{K}{2}} \bigg\},
        \]
        where we use $i_r$ to indicate $r$th element $i$ in the set.
        3) Draw $S^*_{ij} \sim \mbox{Unif}[19,20], \forall (i,j) \in I_{S^*}$.
        4) Lastly, make it symmetric by setting $S^*_{ji}=S^*_{ij},\forall 1\leq i < j \leq n$.
    \item Create an upper-triangular part of the adjacency matrix $X$ whose each entry $X_{ij}$ follows Bernoulli distribution.
        The distribution's parameter is parametrized by a probability, $P^{*}_{ij}=\frac{\exp (\alpha^{*}+f^{*T}_{i}D^{*}f^{*}_{j} +  S^{*}_{ij})}{1+\exp(\alpha^{*}+f^{*T}_{i}D^{*}f^{*}_{j} +  S^{*}_{ij})}$.
        After drawing all the entries of $X$ in the upper-triangular part, then make the matrix symmetric by setting $X_{ji}=X_{ij},\forall 1 \leq i < j \leq n$.
\end{enumerate}

@@

All six networks that are elaborated in scenarios $1$ and $2$ are visualized in Fig.\ref{fig:figure1}.
Notice that the nodes that share the common topics are clustered,
and the cross clustered links are ad-hoc citations.
%ad-hoc links, which connect clusters of papers with $1$ topic, are colored in green.
%Readers can refer the R code provided in $\dots$, for more detailed descriptions on parametric settings for generating networks in Fig\ref{fig:figure1}.

@@

\subsubsection{Choosing the tuning parameters and evaluation criteria} \label{sec:Eval-crit}
Choosing a good pair of tuning parameters is an important yet challenging issue in our setting.
Here we present a heuristic procedure for choosing a good pair of tuning parameters $(\gamma,\delta)$.
Following the scree-plot approach in Ji and Jin \cite{ji2016coauthorship}, we plot the largest $15$ eigenvalues of the adjacency matrix $X$, and find an ``elbow'' point where the eigenvalues seem to level off.
An index of the point, which is to the left of this elbow point, is considered as the number of the topics embedded in the network.
(We will denote this number as $\widehat{K}^{\text{Scree}}$.)
We want to note the readers that the scree-plot analysis serves as a good guideline for determining the range of grids to search over.
With the estimate of the number of topics in the network in mind, we record the $\text{rank}(\widehat{L}^{\gamma,\delta})$ and $|\widehat{S}^{\gamma,\delta}|$ for each tuning parameter pair on a given grid.
We need to go through several iterations of this recording procedure to find a proper range of grid, in which we can get $\text{rank}(\widehat{L}^{\gamma,\delta})=\widehat{K}^{\text{Scree}}$
and $10^{-4}|X| \leq |\widehat{S}^{\gamma,\delta}| \leq 10^{-1}|X|$, via adjusting the range of grid for $\gamma$ and $\delta$ repeatedly.
Once we find a grid, which satisfies above constraints, we choose a pair of tuning parameters: %($\gamma^{\text{Heu}},\delta^{\text{Heu}}$):
\[
    (\gamma^{\text{Heu}},\delta^{\text{Heu}}):=
    \big\{(\gamma,\delta): \text{rank}(\widehat{L}^{\gamma,\delta})=\widehat{K}^{\text{Scree}},
    \mbox{mode} |\widehat{S}^{\gamma,\delta}|: \mbox{subject to} \Hquad 10^{-4}|X| \leq |\widehat{S}^{\gamma,\delta}| \leq 10^{-1}|X| \big\}.
\]
One might wonder how traditional model selection methods, such as the Bayes Information Criterion (BIC;\cite{schwarz1978estimating}) and the AIC, work where we recall that BIC and AIC are defined as follows:
\[
\mbox{BIC}(M) = -2  \mathbb{L}_n( \hat{\beta}(M)) + |M| \log \bigg(\frac{n(n-1)}{2}\bigg),
\]
and
\[
\mbox{AIC}(M) = -2  \mathbb{L}_n( \hat{\beta}(M)) + 2|M|.
\]
Here $M$ indicates the current model, which is implicitly understood that the model is obtained from certain tuning parameter pair $(\gamma,\delta)$.
We use $\mathbb{L}_n( \hat{\beta}(M))$ to denote the maximal log-likelihood for a given model $M$,
and $|M|$ is the number of free parameters in $M$, which is determined by the number of non-zeros in $\widehat{S}^{\gamma,\delta}$ and the low-rank matrix $\widehat{L}^{\gamma,\delta}$.
In detail, if we have $\text{rank}(\widehat{L}^{\gamma,\delta})=K$, we can establish the following
\[
|M| = \sum_{i < j}1_{\{S_{ij} \neq 0\}} + n K - \frac{K(K-1)}{2} + 1 ;
\]
since the number of free parameters in $\widehat{L}^{\gamma,\delta}$ is $K$ plus $nK - K(K+1)/2$, which is the number of free parameters in determining $K$ orth-normal vectors.
Additional $1$ in the last term is due to $\hat{\alpha}$.
We want to find a pair $(\gamma,\delta)$, which minimizes BIC($M$) and AIC($M$) as a function of $(\gamma,\delta)$ respectively, where we denote them as follows:
\[
    (\gamma^{BIC},\delta^{BIC}):= \mbox{arg min}_{\gamma, \delta}
    \mbox{BIC}(M),
    \quad
    (\gamma^{AIC},\delta^{AIC}):= \mbox{arg min}_{\gamma, \delta}
    \mbox{AIC}(M).
\]
We evaluate the models that are selected via our heuristic approach, BIC, and AIC by using following four evaluation metrics:

>>>>

\subsubsection{Synthetic Setting} \label{sec:Syn-set}
%In this subsection, we consider the following $3$ scenarios:
%In this subsection, we consider $3$ scenarios.
%In first scenario, we examine a simple setting, in which every paper in the network has only one topic.
%In second scenario, we study a more complicated situation than the first one where each paper in the network can possibly have more than one topic.
%It is worth noting that different clusters of papers, where each cluster is comprised of papers with same topic, are connected via ad-hoc links.
%In third scenario, we consider a situation
Before specifying the network settings in two scenarios that we want to explore, let us first describe a set of steps for setting the model parameters, $\alpha^*, F^*, D^*$ and $S^*$ sequentially.
We put astroids in the superscripts of parameters to indicate that they are the ground truth.
Readers can refer the meaning of each parameter in the model in Section \ref{sec:intro} and  \ref{sec:model-form}.

@@

    \item Recall that the binary factor loading matrix $F^*$ encodes the relation between topics and papers
        (i.e., if $i$th paper studies about $k$th topic, we denote $F^*_{ki}=1$, otherwise $F^*_{ki}=0$).
        First, we assume that there are $n$ papers in the network, and $K$ topics are embedded in it.
        Each of them consists of roughly $\frac{n}{K}$ papers.
        This can be expressed in $F^*$ as follows:
        \[
            F^* =
               \begin{bmatrix}
                 \smash[b]{\block{n/K}} \\
                 && \smash[b]{\block{n/K}} \\
                 &&&& \ddots \\
                 &&&&& \block{n/K}
                \end{bmatrix}
              \in R^{K \times n},
        \]
        where each row of $F^*$ has $\frac{n}{K}$ $1$'s and each column has only one $1$.
        Note that the remaining entries of the matrix are filled with zeros.
        Then, we assume that $n_{l}$ papers share $l$ topics and $n_{m}$ papers share $m$ topics ($1 < l < m \leq K$).
        This can be represented in the $F^*$ in following steps:
        $1)$ Pick distinct $n_{l}$ indices randomly from $\{1,2,\dots,n\}$.
        We will denote the set of the indices as $\Omega_{n_l}$.
        $2)$ Choose $n_{m}$ indices from the set $\{1,2,\dots,n\}\setminus \Omega_{n_{l}}$ and denote the set of those indices as $\Omega_{n_{m}}$.
        $3)$ Make the columns of $F^*$ with corresponding indices in set $\Omega_{n_{l}} \cup \Omega_{n_{m}}$ zeros.
        $4)$ We use a notation $f^*_j$ to denote the $j$th column of the matrix $F^*$.
        Fill arbitrary $l$ entries of $f^*_j$ for $j \in \Omega_{n_l}$ with $1$s, and also fill arbitrary $m$ entries of $f^*_j$ for $j \in \Omega_{n_m}$ with $1$s.
        Lastly, we set $F^*=JF^*$ where $J=I_{n}-\frac{1}{n}\mathbbm{1}\mathbbm{1}^T$.

@@

    \item Recall that the positive entries of $S^*$ can characterize the links in the network, which cannot be accounted by the common topics.
        In the spirit of this notion, we construct ad-hoc links in a way that they connect the clusters of papers with one topic.
        Specifically, we assume that there are $\frac{|S^*|}{\binom{K}{2}}$ edges between two clusters, where $|S^*|$ denotes the number of non-zero entries of the upper-triangular part of the matrix $S^*$.
        This can be implemented via the following steps:
        1) We construct $K$ sets $C^*_1,\dots,C^*_K$ that are defined as follows:
        \begin{align*}
           &C^*_{1} \subseteq \big\{1, 2, \dots, \frac{n}{K} \big\} \setminus \big\{ \Omega_{n_l}
           \cup \Omega_{n_m} \big\} \\
           &C^*_{2} \subseteq \big\{\frac{n}{K}+1, \dots, \frac{2n}{K} \big\} \setminus \big\{ \Omega_{n_l} \cup \Omega_{n_m} \big\}  \\
           & \quad \vdots \\
           &C^*_{K} \subseteq \big\{\frac{(K-1)n}{K}+1, \dots, n \big\} \setminus \big\{ \Omega_{n_l}
           \cup \Omega_{n_m} \big\}
        \end{align*}
        where each of them has arbitrary $\frac{|S^*|}{\binom{K}{2}}$ elements.
        2) Create a set $I_{S^*}$ whose elements are pairs of indices such that
        \[
            I_{S^*} =
            \bigg\{ (i_r,j_r) : i_r \in C^*_p, j_r \in C^*_q, 1 \leq p < q \leq K,   r = 1,2,\dots,\frac{|S^*|}{\binom{K}{2}} \bigg\},
        \]
        where we use $i_r$ to indicate $r$th element $i$ in the set.
        3) Draw $S^*_{ij} \sim \mbox{Unif}[19,20], \forall (i,j) \in I_{S^*}$.
        4) Lastly, make it symmetric by setting $S^*_{ji}=S^*_{ij},\forall 1\leq i < j \leq n$.
    \item Create an upper-triangular part of the adjacency matrix $X$ whose each entry $X_{ij}$ follows Bernoulli distribution.
        The distribution's parameter is parametrized by a probability, $P^{*}_{ij}=\frac{\exp (\alpha^{*}+f^{*T}_{i}D^{*}f^{*}_{j} +  S^{*}_{ij})}{1+\exp(\alpha^{*}+f^{*T}_{i}D^{*}f^{*}_{j} +  S^{*}_{ij})}$.
        After drawing all the entries of $X$ in the upper-triangular part, then make the matrix symmetric by setting $X_{ji}=X_{ij},\forall 1 \leq i < j \leq n$.
\end{enumerate}

@@

All six networks that are elaborated in scenarios $1$ and $2$ are visualized in Fig.\ref{fig:figure1}.
Notice that the nodes that share the common topics are clustered,
and the cross clustered links are the ad-hoc citations.
%ad-hoc links, which connect clusters of papers with $1$ topic, are colored in green.
%Readers can refer the R code provided in $\dots$, for more detailed descriptions on parametric settings for generating networks in Fig\ref{fig:figure1}.

@@

\subsubsection{Choosing the tuning parameters and evaluation criteria} \label{sec:Eval-crit}
Choosing a good pair of tuning parameters is an important yet challenging issue in our setting.
Here we present a heuristic procedure for choosing a good pair of tuning parameters $(\gamma,\delta)$.
Following the scree-plot approach in Ji and Jin \cite{ji2016coauthorship}, we plot the largest $15$ eigenvalues of the adjacency matrix $X$, and find an ``elbow'' point where the eigenvalues seem to level off.
An index of the point, which is to the left of this elbow point, is considered as the number of the topics embedded in the network.
(We will denote this number as $\widehat{K}^{\text{Scree}}$.)
We want to note that the scree-plot analysis serves as a good guideline for determining the range of grids to search over.
With the estimate of the number of topics in the network in mind, we record the $\text{rank}(\widehat{L}^{\gamma,\delta})$ and $|\widehat{S}^{\gamma,\delta}|$ for each tuning parameter pair on a given grid.
We need to go through several iterations of this recording procedure to find a proper range of grid, in which we can get $\text{rank}(\widehat{L}^{\gamma,\delta})=\widehat{K}^{\text{Scree}}$
and $10^{-4}|X| \leq |\widehat{S}^{\gamma,\delta}| \leq 10^{-1}|X|$, via adjusting the range of grid for $\gamma$ and $\delta$ repeatedly.
Once we find a grid, which satisfies above constraints, we choose a pair of tuning parameters: %($\gamma^{\text{Heu}},\delta^{\text{Heu}}$):
\[
    (\gamma^{\text{Heu}},\delta^{\text{Heu}}):=
    \big\{(\gamma,\delta): \text{rank}(\widehat{L}^{\gamma,\delta})=\widehat{K}^{\text{Scree}},
    \mbox{mode} |\widehat{S}^{\gamma,\delta}|: \mbox{subject to} \Hquad 10^{-4}|X| \leq |\widehat{S}^{\gamma,\delta}| \leq 10^{-1}|X| \big\}.
\]
One might wonder how traditional model selection methods, such as the Bayes Information Criterion (BIC;\cite{schwarz1978estimating}) and the AIC, work where we recall that BIC and AIC are defined as follows:
\[
\mbox{BIC}(M) = -2  \mathbb{L}_n( \hat{\beta}(M)) + |M| \log \bigg(\frac{n(n-1)}{2}\bigg),
\]
and
\[
\mbox{AIC}(M) = -2  \mathbb{L}_n( \hat{\beta}(M)) + 2|M|.
\]
Here $M$ indicates the current model, which is implicitly understood that the model is obtained from certain tuning parameter pair $(\gamma,\delta)$.
We use $\mathbb{L}_n( \hat{\beta}(M))$ to denote the maximal log-likelihood for a given model $M$,
and $|M|$ is the number of free parameters in $M$, which is determined by the number of non-zeros in $\widehat{S}^{\gamma,\delta}$ and the low-rank matrix $\widehat{L}^{\gamma,\delta}$.
In detail, if we have $\text{rank}(\widehat{L}^{\gamma,\delta})=K$, we can establish the following
\[
|M| = \sum_{i < j}1_{\{S_{ij} \neq 0\}} + n K - \frac{K(K-1)}{2} + 1 ;
\]
since the number of free parameters in $\widehat{L}^{\gamma,\delta}$ is $K$ plus $nK - K(K+1)/2$, which is the number of free parameters in determining $K$ orth-normal vectors.
Additional $1$ in the last term is due to $\hat{\alpha}$.
We want to find a pair $(\gamma,\delta)$, which minimizes BIC($M$) and AIC($M$) as a function of $(\gamma,\delta)$ respectively, where we denote them as follows:
\[
    (\gamma^{BIC},\delta^{BIC}):= \mbox{arg min}_{\gamma, \delta}
    \mbox{BIC}(M),
    \quad
    (\gamma^{AIC},\delta^{AIC}):= \mbox{arg min}_{\gamma, \delta}
    \mbox{AIC}(M).
\]
We evaluate the models that are selected via our heuristic approach, BIC, and AIC by using following four evaluation metrics:
